\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-28 paper.qmd
lots of structuring

@@

Projections of the $7\text{-}D$ S-curve data (purple points) overlaid with the model (black grid with green points indicating bin means) constructed from the UMAP layout, with 60 non-empty bins (using the **langevitour** software). The model closely fits the shape of this data. A video of the full tour is available at (<https://youtu.be/G1m4q9k--v4>).
:::

@@

#### Fitted values

The fitted values of the points in $H_h$ refers to the \pD{} mapping $C_{h}^{(p)}$ of the corresponding \kD{} model point $C_{h}^{(2)}$.

#### Residuals and error

To assess how well our method captures and represents the underlying structure of the \pD{} data, residuals are important. Residuals are computed by taking the squared \pD{} Euclidean distance, $\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{P} (x_{hij} - C^{(p)}_{hj})^2$. Additionally, Mean Squared Error (MSE), $\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{P} (x_{hij} - C^{(p)}_{hj})^2$, is computed based on residuals.

### Prediction

The prediction approach involves finding the nearest \kD{} model point for a new \pD{} point. We define the function $z: \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{m\times p}$, where $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{P}(x_{iv} - C^{(p)}_{jv})^2}$ maps each \pD{} point to its nearest \pD{} mapping of the model. Therefore, the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$ be the predicted values.

@@

```{r}
#| echo: false
#| label: fig-lwd-scurve
#| fig-pos: H
#| fig-cap: "Two approaches to identify low-density hexagons. Hexagon grid (a) marks hexagons with a standardized count less than $0.25$. Hexagon grid (b) marks hexagons by considering the densities of neighboring bins. Removing hexagons by also investigating the densities of neighboring bins is more reliable for preserving the structure."
#| fig-width: 9
#| fig-height: 6

hex_grid_rm_lwd_scurve1 + hex_grid_rm_lwd_scurve2 +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

## Deciding on the most suitable fit

@@

@@

## Illustration on simulated data

>>>>

XXX SHOW THE S-CURVE DATA WITH THE TRUE MODEL OVERLAID HERE, BEFORE THE NLDR VIEWS.

@@

XXX Model in 2D, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has breaks which means that it does not adequately capture the surface. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/G1m4q9k--v4>). NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.
:::

@@

The model here is similar to a confirmatory factor analysis model (REF), $\widehat{T}(X_1, X_2, X_3) + \Epsilon$. The difference between the fitted model and observed values would be considered to be residuals, and for this problem are $7\text{-}D$.

<!--#### Fitted values-->

Observations are associated with their bin center, $C_{h}^{(p)}$, which are also considered to be the *fitted values*. These can also be denoted as $\widehat{X}$. <!--The fitted values of the points in $H_h$ refers to the \pD{} mapping $C_{h}^{(p)}$ of the corresponding \kD{} model point $C_{h}^{(2)}$.-->

<!--#### Error-->

The error is computed by taking the squared \pD{} Euclidean distance, corresponding to computing the mean squared error (MSE) as:

$$\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2$$

where $n$ is XXX, $b$ is XXX, $n_h$ is XXX, $p$ is XXX, $\mathbfit{x}_{hij}$ is XXX. MAKE THIS A NUMBERED EQUATION, SO WE CAN REFER TO IT LATER

### Prediction into \gD{}

A new benefit of this fitted model is that it allows us to now predict a new observation's value in the NLDR, for any method. The steps are to determine the closest bin center in \pD{}, $C^{(p)}_{h}$ and predict it to be the centroid of this bin in \gD{}, $C^{(2)}_{h}$. This can be written as, let $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$, then the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$.

<!--The prediction approach involves finding the nearest \kD{} model point for a new \pD{} point. We define the function $z: \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{m\times p}$, where $z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}$ maps each \pD{} point to its nearest \pD{} mapping of the model. Therefore, the new observation $i$ falls in the hexagon, $H_h = \{i| z(i) = h\}$ and the corresponding \kD{} bin centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$ be the predicted values.-->

@@

<!--
```{r}
#| echo: false
#| label: fig-lwd-scurve
#| fig-pos: H
#| fig-cap: "Two approaches to identify low-density hexagons. Hexagon grid (a) marks hexagons with a standardized count less than $0.25$. Hexagon grid (b) marks hexagons by considering the densities of neighboring bins. Removing hexagons by also investigating the densities of neighboring bins is more reliable for preserving the structure."
#| fig-width: 9
#| fig-height: 6

hex_grid_rm_lwd_scurve1 + hex_grid_rm_lwd_scurve2 +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

## Best fit

Deciding on the best fit relies on several elements:

- the choice of NLDR method, and the parameters used to create it, and
- model fit parameters: bin size, low density bin removal, long edge removal.

Comparing the MSE to obtain the best fit is suitable if one starts from the same NLDR representation. In theory, because the MSE is computed on \pD{} measuring the fit between model and data it might still be useful to compare different NLDR representations. A good NLDR representation should produce a good fit, producing a low MSE if the model fits the data well. However, it technically might be quite variable.

@@

XXX BEST FIT FOR S-CURVE MIGHT BE A DIFFERENT NLDR. 
SHOW THE SEVERAL SUB-OPTIMAL FITS.

@@

## A curious difference between t-SNE and UMAP revealer

XXX THIS IS NOT INTERESTING IN ITS CURRENT FORM. THIS IS ONLY ANY INTERESTING EXAMPLE WHEN THE FILLED OUT VS FLAT SHAPES ARE DISCUSSED.

### 2024-06-28 paper.tex
small changes to headers
lots of structuring

\caption{\label{fig-scurve-sc}Projections of the \(7\text{-}D\) S-curve
data (purple points) overlaid with the model (black grid with green
points indicating bin means) constructed from the UMAP layout, with 60
non-empty bins (using the \textbf{langevitour} software). The model
closely fits the shape of this data. A video of the full tour is
available at (\url{https://youtu.be/G1m4q9k--v4}).}

@@

\subsubsection{Fitted values}\label{fitted-values}

The fitted values of the points in \(H_h\) refers to the \pD{} mapping
\(C_{h}^{(p)}\) of the corresponding \kD{} model point \(C_{h}^{(2)}\).

\subsubsection{Error calculation}\label{error-calculation}

To assess how well our method captures and represents the underlying
structure of the \pD{} data, residuals are important. Residuals are
computed by taking the squared \pD{} Euclidean distance,
\(\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{P} (x_{hij} - C^{(p)}_{hj})^2\).
Additionally, Mean Squared Error (MSE),
\(\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{P} (x_{hij} - C^{(p)}_{hj})^2\),
is computed based on residuals.

\subsection{Prediction}\label{prediction}

The prediction approach involves finding the nearest \kD{} model point
for a new \pD{} point. We define the function
\(z: \mathbb{R}^{n\times p} \rightarrow \mathbb{R}^{m\times p}\), where
\(z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{P}(x_{iv} - C^{(p)}_{jv})^2}\)
maps each \pD{} point to its nearest \pD{} mapping of the model.
Therefore, the new observation \(i\) falls in the hexagon,
\(H_h = \{i| z(i) = h\}\) and the corresponding \kD{} bin centroids,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\) be the predicted values.

@@

Once setting up the hexagon grid with an appropriate number of bins,
some hexagon bins may have few or no data points within them (see
Figure~\ref{fig-bins-scurve} (b)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagon bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid. Next, bins with a standard count less than a
benchmark value are removed (see Figure~\ref{fig-lwd-scurve} (a)). There
is no specific rule for selecting a benchmark value. However, the
following steps can help determine a suitable value for removing
low-density hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven \gD{} model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density bins (see
Figure~\ref{fig-lwd-scurve} (b)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the \gD{} model.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-lwd-scurve-1.pdf}

\caption{\label{fig-lwd-scurve}Two approaches to identify low-density
hexagons. Hexagon grid (a) marks hexagons with a standardized count less
than \(0.25\). Hexagon grid (b) marks hexagons by considering the
densities of neighboring bins. Removing hexagons by also investigating
the densities of neighboring bins is more reliable for preserving the
structure.}

@@

\section{Deciding on the most suitable
fit}\label{deciding-on-the-most-suitable-fit}

@@

\section{Illustration on simulated
data}\label{illustration-on-simulated-data}

@@

\subsection{pbmc}\label{pbmc}

@@

\subsection{digits: 1}\label{digits-1}

>>>>

\caption{\label{fig-scurve-sc}XXX Model in 2D, on the UMAP layout, and
three views of the fit in projections from \(7\text{-}D\), for the
S-curve data. The model closely fits the shape, but it has breaks which
means that it does not adequately capture the surface. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at (\url{https://youtu.be/G1m4q9k--v4}).
NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.}

@@

The model here is similar to a confirmatory factor analysis model (REF),
\(\widehat{T}(X_1, X_2, X_3) + \Epsilon\). The difference between the
fitted model and observed values would be considered to be residuals,
and for this problem are \(7\text{-}D\).

Observations are associated with their bin center, \(C_{h}^{(p)}\),
which are also considered to be the \emph{fitted values}. These can also
be denoted as \(\widehat{X}\).

The error is computed by taking the squared \pD{} Euclidean distance,
corresponding to computing the mean squared error (MSE) as:

\[\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2\]

where \(n\) is XXX, \(b\) is XXX, \(n_h\) is XXX, \(p\) is XXX,
\(\mathbfit{x}_{hij}\) is XXX. MAKE THIS A NUMBERED EQUATION, SO WE CAN
REFER TO IT LATER

\subsection{\texorpdfstring{Prediction into
\gD{}}{Prediction into }}\label{prediction-into}

A new benefit of this fitted model is that it allows us to now predict a
new observation's value in the NLDR, for any method. The steps are to
determine the closest bin center in \pD{}, \(C^{(p)}_{h}\) and predict
it to be the centroid of this bin in \gD{}, \(C^{(2)}_{h}\). This can be
written as, let
\(z(i) = \arg\min_{j = 1, \dots, b} \sqrt{\sum_{v=1}^{p}(x_{iv} - C^{(p)}_{jv})^2}\),
then the new observation \(i\) falls in the hexagon,
\(H_h = \{i| z(i) = h\}\) and the corresponding \kD{} bin centroids,
\(C_{h}^{(2)} = (c_{h1}, c_{h2})\).

@@

Once setting up the hexagon grid with an appropriate number of bins,
some hexagon bins may have few or no data points within them (see
Figure~\ref{fig-bins-scurve} (b)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagon bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid. Next, bins with a standard count less than a
benchmark value are removed (see \textbf{?@fig-lwd-scurve} (a)). There
is no specific rule for selecting a benchmark value. However, the
following steps can help determine a suitable value for removing
low-density hexagons:

@@

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven \gD{} model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density bins (see
\textbf{?@fig-lwd-scurve} (b)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the \gD{} model.

@@

\section{Best fit}\label{best-fit}

Deciding on the best fit relies on several elements:

\begin{itemize}
\tightlist
\item
  the choice of NLDR method, and the parameters used to create it, and
\item
  model fit parameters: bin size, low density bin removal, long edge
  removal.
\end{itemize}

Comparing the MSE to obtain the best fit is suitable if one starts from
the same NLDR representation. In theory, because the MSE is computed on
\pD{} measuring the fit between model and data it might still be useful
to compare different NLDR representations. A good NLDR representation
should produce a good fit, producing a low MSE if the model fits the
data well. However, it technically might be quite variable.

@@

XXX BEST FIT FOR S-CURVE MIGHT BE A DIFFERENT NLDR. SHOW THE SEVERAL
SUB-OPTIMAL FITS.

\section{A curious difference between t-SNE and UMAP
revealer}\label{a-curious-difference-between-t-sne-and-umap-revealer}

XXX THIS IS NOT INTERESTING IN ITS CURRENT FORM. THIS IS ONLY ANY
INTERESTING EXAMPLE WHEN THE FILLED OUT VS FLAT SHAPES ARE DISCUSSED.

@@

\subsection{Single-cell gene
expression}\label{single-cell-gene-expression}

@@

\subsection{Hand-written digits}\label{hand-written-digits}
